openTSNE: a modular \proglang{Python} library for t-SNE dimensionality reduction and embedding

### 2023-06-02 2023-05-revision-1/main.tex
minor changes

We use the Kullback-Leibler (KL) divergence to measure the agreement
between the two distributions $\mathbf{P}$ and $\mathbf{Q}$
\begin{equation}
C = \text{KL}(\mathbf{P} \mid \mid \mathbf{Q}) = \sum_{ij} p_{ij} \log \frac{p_{ij}}{q_{ij}}.
\label{eq:kl_divergence}
\end{equation}
The objective is to find embeddings $\mathbf{Y}$ that minimize the KL divergence. The corresponding gradient takes the form
\begin{equation}
\frac{\partial C}{\partial \mathbf{y}_i} = 4 \sum_{j \neq i} \left ( p_{ij} - q_{ij} \right ) \left ( \mathbf{y}_i - \mathbf{y}_j \right ) w_{ij},
\label{eq:tsne_gradient}
\end{equation}
where $w_{ij} = \left ( 1 + || \mathbf{y}_i - \mathbf{y}_j || ^2 \right )^{-1}$
and represents the unnormalized $q_{ij}$.
Alternatively, this gradient can be rewritten as
\begin{equation}
\frac{\partial C}{\partial \mathbf{y}_i} = 4 \Bigg [
\underbrace{\sum_{j \neq i} p_{ij} q_{ij} Z \left ( \mathbf{y}_i - \mathbf{y}_j \right )}_{\text{attractive forces}}  -
\underbrace{\sum_{j \neq i} q_{ij}^2 Z \left ( \mathbf{y}_i - \mathbf{y}_j \right )}_{\text{repulsive forces}}
\Bigg ], \label{eq:grad_attr_rep}
\end{equation}
where $Z = \sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2 \right )^{-1}$. This formulation implies that the t-SNE algorithm can be interpreted as a force-directed layout algorithm, where individual data points act as particles exerting both attractive and repulsive on each other until a state of equilibrium is reached.

@@

Similarly to standard t-SNE, a direct calculation of gradients takes $\mathcal{O}(N \cdot M)$ time, but it is straightforward to adapt the Barnes-Hut and FIt-SNE approximation schemes, reducing the time complexity to $\mathcal{O}(M \log N)$ and $\mathcal{O}(M)$, respectively. Special care must be taken to adjust the learning rate during optimization as the parameter values used during standard optimization lead to highly unstable behaviour. Empirically, we have found that reducing the learning rate to $\eta \sim 0.1$ results in more stable optimization, as discussed in \citet{policar2021embedding}. Moreover, during standard optimization, data points are positioned in the embedding such that an equilibrium between the attractive and repulsive forces is achieved. However, when embedding new data points, the reference embedding remains fixed. Consequently, if a new data point needs to be positioned between existing reference data points, it may not be possible to reach such an equilibrium, as the reference data points cannot adjust to accomodate the new data point. To compensate for the increased repulsive forces, we have found it beneficial to increase the exaggeration factor to $\rho=1.5$ during optimization. The training dynamics of embedding new data points into existing embeddings are yet fully understood and are left to future exploration.

@@

>>>>

We use the Kullback-Leibler (KL) divergence to measure the agreement
between the two distributions $\mathbf{P}$ and $\mathbf{Q}$
\begin{equation}
C = \text{KL}(\mathbf{P} \mid \mid \mathbf{Q}) = \sum_{ij} p_{ij} \log \frac{p_{ij}}{q_{ij}}.
\label{eq:kl_divergence}
\end{equation}
The objective is to find embeddings $\mathbf{Y}$ that minimize the KL divergence. The corresponding gradient takes the form
\begin{equation}
\frac{\partial C}{\partial \mathbf{y}_i} = 4 \sum_{j \neq i} \left ( p_{ij} - q_{ij} \right ) \left ( \mathbf{y}_i - \mathbf{y}_j \right ) w_{ij},
\label{eq:tsne_gradient}
\end{equation}
where $w_{ij} = \left ( 1 + || \mathbf{y}_i - \mathbf{y}_j || ^2 \right )^{-1}$
and represents the unnormalized $q_{ij}$.
We can rewrite this gradient to
\begin{equation}
\frac{\partial C}{\partial \mathbf{y}_i} = 4 \Bigg [
\underbrace{\sum_{j \neq i} p_{ij} q_{ij} Z \left ( \mathbf{y}_i - \mathbf{y}_j \right )}_{\text{attractive forces}}  -
\underbrace{\sum_{j \neq i} q_{ij}^2 Z \left ( \mathbf{y}_i - \mathbf{y}_j \right )}_{\text{repulsive forces}}
\Bigg ], \label{eq:grad_attr_rep}
\end{equation}
where $Z = \sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2 \right )^{-1}$. With this formulation, we can interpret t-SNE algorithm as a force-directed layout algorithm, where individual data points act as particles exerting both attractive and repulsive on each other until a state of equilibrium is reached.

@@

Similarly to standard t-SNE, a direct calculation of gradients takes $\mathcal{O}(N \cdot M)$ time, but it is straightforward to adapt the Barnes-Hut and FIt-SNE approximation schemes, reducing the time complexity to $\mathcal{O}(M \log N)$ and $\mathcal{O}(M)$, respectively. Special care must be taken to adjust the learning rate during optimization as the parameter values used during standard optimization may lead to  unstable behaviour. We have observed that reducing the learning rate to $\eta \sim 0.1$ results in more stable optimization, as discussed in \citet{policar2021embedding}. Standard optimization positions data points in the embedding to achieve an equilibrium between the attractive and repulsive forces. However, when embedding new data points, the reference embedding remains fixed. Consequently, if a new data point needs to be positioned among existing reference data points, it may not be possible to reach such an equilibrium, as the reference data points cannot adjust to accomodate the new data point. To compensate for the increased repulsive forces, it is beneficial to increase the exaggeration factor to $\rho=1.5$ during optimization.

@@

\subsubsection{Comparison to UMAP}

We compare the performance of the \opentsne (v1.0.0) and \pkg{umap-learn} (0.5.3) libraries. \pkg{umap-learn} is the official \textsf{Python} implementation of the UMAP algorithm~\citep{2018arXivUMAP}, an alternative dimensionality reduction technique used for the visualization of high-dimensional data. A throughout comparison of the two algorithms is outside the scope of this manuscript, and we here instead highlight only their similarities and differences as they relate to the runtime of the implementations.

Both t-SNE and UMAP are force-directed layout algorithms in which each data point acts as a particle, attracting and repelling other particles. In the first step, both algorithms use the high-dimensional data set to construct a $k$-nearest neighbors graph (KNNG). Then, the low-dimensional embedding is optimized such that data points connected in the KNNG attract one another, and all data points exert some small, repulsive force. The two methods differ in the KNNG construction process, the attractive and repulsive forces specification, and their optimization approaches.

The primary factor dictating the runtime of both methods is the number of nearest neighbors used when constructing the KNNG. Apart from the runtime required to find the $k$-nearest neighbors, the number of neighbors directly influences the runtime of each optimization step. In each step, every data point exerts an attractive force on each neighbor in the low-dimensional embedding. Therefore, the larger the number of neighbors considered, the longer the runtime. Modern implementations of the t-SNE algorithm construct a KNNG with $3 \cdot u$ nearest neighbors where $u$ is the desired perplexity. By default, \opentsne (v1.0.0) uses a perplexity value of 30, meaning that 90 nearest neighbors must be found and considered for each data point. In contrast, the \pkg{umap-learn} package (v0.5.3) constructs a KNNG using only 15 neighbors by default, substantially reducing the time spent on nearest neighbor search and point interactions compared to t-SNE.

Alternatively, empirical evidence indicates that replacing the standard perplexity-based Gaussian affinity model in t-SNE with a uniform affinity model produces markedly similar embeddings to that of standard t-SNE. For instance, running standard t-SNE with the perplexity of $u=30$ constructs a KNNG with $3 \cdot u = 90$ neighbors. However, when using the equivalent uniform affinity model, only $u$ neighbors are needed, making the uniform affinity model attractive for visualizing large data sets.

In Fig.~\ref{fig:benchmarks_umap}, we include benchmarks of the standard perplexity-based t-SNE algorithm, its uniform affinity model variant, and UMAP. Differently from the previous benchmarks, which were run using older parameter settings to accommodate different t-SNE implementations, we now run both libraries using their default parameter settings to measure the runtime of typical, real-world use cases.
\begin{figure*}[ht]
  \centering
  \includegraphics{benchmarks_umap-final}
  \caption{\label{fig:benchmarks_umap} We benchmark \opentsne (v1.0.0) against \pkg{umap-learn} (v0.5.3), the official \textsf{Python} implementation of the UMAP algorithm. Notice that the standard single-threaded perplexity-based variant of t-SNE runs slower than UMAP but outperforms UMAP in the multi-threaded case. We also benchmark the uniform affinity model variant of t-SNE, which outperforms both standard t-SNE and UMAP.}
\end{figure*}

Given that, by default, UMAP considers six times fewer neighbors than standard t-SNE, it is unsurprising that the single-threaded variant of UMAP outperforms the single-threaded implementation of standard t-SNE. Surprisingly, however, the single-threaded uniform affinity variant of t-SNE outperforms the single-threaded variant of UMAP and its multi-threaded variant, despite dealing with half as many neighbors. This result likely stems from the fact that UMAP is written in pure \textsf{Python} but uses \pkg{numba}~\citep{lam2015numba} to accelerate computation. On the other hand, we wrote critical elements of the \opentsne library in \pkg{cython}~\citep{behnel2011cython}, which is subsequently transpiled into \textsf{C++}, which may be able to be more aggressively optimized by the \textsf{C++} compiler. Another contributing factor may be the efficiency of the approximation schemes. While the UMAP and FIt-SNE algorithms are asymptotically linear in the number of samples, their real-world runtimes may differ significantly.

The multi-threaded benchmarks show that \opentsne outperforms UMAP in standard and uniform variants. We can attribute this to different optimization procedures used by the two algorithms. At each optimization step, \opentsne computes all pairwise interactions between data points in the embedding. While directly calculating these interactions would result in quadratic time complexity, \opentsne implements efficient approximation schemes that reduce this to linear time complexity (see Sec.~\ref{sec:meth.approx}). Importantly, these approximation schemes are very amenable to parallelization. The runtime differences between the single-core and multi-core performance of \opentsne clearly demonstrate this. On the other hand, UMAP bypasses the need to compute all pairwise interactions between data points using negative sampling, allowing it to scale linearly in the number of samples without additional approximation schemes. However, negative sampling is less amenable to parallelization and does not benefit from the availability of additional cores as much as t-SNE.

Lastly, we focus on an often overlooked runtime metric -- the runtime per optimization step. Both t-SNE and UMAP comprise two steps. First, both methods construct a KNNG, which runtime is dominated by $k$-nearest neighbor search. In the second step, a low-dimensional embedding is found via optimization. By default, \opentsne runs optimization for 750 iterations. UMAP, on the other hand, uses a heuristic based on the data set size to determine the number of iterations. For data sets containing fewer than 10,000 data points, UMAP opts for 500 iterations. For larger data sets, however, this is reduced to 200 iterations. Fig.~\ref{fig:benchmarks_umap} indicates that even in the single-threaded scenario, the perplexity-based variant of \opentsne requires significantly less time per iteration compared to \pkg{umap-learn}, as each \opentsne benchmark runs $3.75$ times more iterations than \pkg{umap-learn} despite considering more neighbors at each optimization step.
